Fast Algorithms for Intersection of Non-matching Grids Using \plucker coordinates.

### 2016-09-30 paper_intersections.tex
Improve benchmarks description.

\section{Element Intersections}
\label{sec:element intersecitons}
In this section, we present algorithms for computing intersection of a pair of simplicial elements of a~different dimension in the 3D ambient space. 
In particular we address intersection algorithms for 1D-2D, 1D-3D, 2D-3D pairs of elements. We have implemented the case 2D-2D as well however the treatment 
of the special cases is quite technical and not fully completed yet. The fundamental idea is to compute intersection of 1D-2D simplices using the 
\plucker coordinates and reduce all other cases to this one.

@@

%\subsection{Intersection Triangle-Tetrahedron (2D-2D)}
%\noteJB{Just note that this algorithm can also be implemented using 1d-2d, but quite complex to not yet fully tested}

@@

\subsection{Intersection Triangle-Tetrahedron (2D-3D)}
The intersection of a triangle $S_2$ and a tetrahedron $S_3$ is an $n$-side intersection polygon (IP), $n\le 7$. The sides of the polygon
lie either on sides of $S_2$ or on faces of $S_3$. Thus each vertex (IC) of the polygon
can arise either from side-face intersection, or from edge-triangle intersection, or be a vertex of $S_2$.
To get all ICs, we have to compute at most 12 side-face intersections and at most 6 edge-triangle intersections. However,
to this end we only need to compute 9 \plucker coordinates (3 sides, 6 edges) and 18 permuted inner produts, one for every side-edge pair.
Computation of IP consists of three stages: calculation of side-tetrahedron ICs (Algorithm \ref{algo:colect_23_ip_triangle}), 
calculation of edge-triangle ICs (Algorithm \ref{algo:collect_23_ip_edges}), oredering of ICs (Algorithm \ref{algo:order_23}).
The intersection corners appended to the list $I$ as they are computed however their order on the polygon boundary is 
defined by the \emph{connection tables} $F_g(:)$ and $F_p(:)$. Every side of the polygon that lies on \nface $x\in M_2\cup M_3$ is followed by an IC given by $F_g[x]$ 
and every IC $p$ is followed by the side that lies on $F_p[p]\in M_2\cup M_3$ (see \noteJB{possible figure}). The vertices of the polygon are numbered in the same 
way as is the order and orientation of the sides of $S_2$, that is counterclockwise around the interiror with normal pointing to us.

@@

  \DontPrintSemicolon
  \SetKw{continue}{continue}
  \KwIn{input data}
  \KwOut{List of ICs on sorted output data}  
  $F_g(:)=-1,\ F_p(:)=-1$\tcp*{Unset links.} 
  \For{side $s$ of $S_2$}{ \nllabel{line:one side loop}
    $L$ = intersection($s, S_3$)\;
    \lIf{ $\abs{L} = 0$}{\continue}
    \lIf{ $\abs{L} = 1$}{
        append $p$ to $J$
        \continue
    }
    \For{$p$ in $L$}{
%        \lIf{$p \ne I[-1]$ }{append $p$ to $I$}
        \nllabel{line:m4_object}
        $p$ lies on $x\in M_3$\;
        \lIf{$p$ lies at vertex $v$ of $S_3$}{$x=v$}
        \uIf{$p$ is first in $L$}{
            $F_g[x] = p$,\ $F_p[p] = s$\;
        }
        \Else($p$ is the last in $L$){
            $F_g[s] = p,\ F_p[p] = x$\;
            \lIf{$F_g[x] = -1$}{$F_g[x] = p$} \nllabel{line:back_link43}
        }
    }
  }  
%  \If{$I[-1] = I[0]$}{
%    $x = F_p[I[-1]]$\;
%    \If{$F_g[x]]=I[-1]$}{$F_g[x]=I[0]$}
%    $F_g[s_2] = I[0]$\;
%    remove $I[-1]$
%  }
\end{algorithm}

Algorithm \ref{algo:colect_23_ip_triangle} computes all ICs on the boundary of $S_2$.  It passes through every side $s$ of the triangle $S_2$ 
and computes the line-tetrahedron intersection $L$.
If the side is just touching $S_3$ and we get $L$ with single IC. These ICs will be rediscovered again in Algorithm \ref{algo:collect_23_ip_edges} 
with better topological information, however this is not the case if the touched edge $e$ of $S_3$ is coplanar with $S_2$ and the IC is inside of $e$.
To this end we save the IC into separate list $J$ and skip filling of the connection tables.
In the regular case, we process each of the two ICs in $L$ (loop on line \ref{line:one side loop}). The IC $p$ is added to the list $I$,
vertices of $S_2$ added twice are merged at final stage. Then (line \ref{line:m4_object}) we identify the \nface $x\in M_3$ 
the point $p$ lies on and set the tables $F_g$, $F_p$. For the IC at the vertex of $S_2$ we set $x$ to that vertex regardless of its position on $S_3$.
The backward temporary link on the line \ref{line:back_link43} is used later in Algorithm \ref{algo:collect_23_ip_edges} to allow proper continuation 
of the IC if it lies on edge or at vertex of $S_3$. The condition at the same line deals with the case that two sides of $S_3$ intersect 
the same object $x\in M_3$, in such case the back-link is unnecessary and would overwrite the correct link set by previous IC.

@@

  \lFor{edge $e$ of $S_3$}{$L[e]$ = intersection($e, S_2$)}  \nllabel{line:12edges}
  \For{edge $e$ of $S_3$ with non-empty $L[e]$}{
    $p=L[e]$\;
    \uIf{ $p$ is inside $e$}{ 
        \nllabel{line:edge_faces}        
        $(f_0,\ f_1)$ = \edgefaces($e$)
    } 
    \Else($p$ at the vertex $v$ of $S_3$){       
            $(f_0,\ f_1)$ = \vertexfaces($v$,$L$)   \tcp*{Algorithm \ref{algo:vertex_faces}}
            \nllabel{line:mark_edges}
    }    
    \uIf{ $p$ is on boundary of $S_2$ }{
         $p$ lies on edge or at vertex $x \in M_3$\;
         $q=F_g[x]$\tcp*{$q$ is already computed $p$}
         \lIf{$F_p[q]=x$}{$F_p[q]=f_1$}
         \lElse{$F_g[f_0]=q$}
         $F_g[x]=-1$\tcp*{remove the backlink}
    }
    \Else{
         append $p$ to $I$\;
         $F_g[f_0]=p,\ F_p[p]=f_1$\tcp*{overwrite the backlink}
    }   
  }  
\end{algorithm}

Algorithm \ref{algo:collect_23_ip_edges} uses the line-triangle intersection algorithm for the edges of $S_3$ (line \ref{line:12edges}).
First, the intersection $L[e]$ is evaluated for every edge $e$. The loop produces ICs in the interior of $S_2$ and possibly those ICs with special position on vertex or edge of $S_3$ already computed 
in Algorithm \ref{algo:colect_23_ip_triangle}. Then we pass through once again skipping edges with none or degenerate $L[e]$.
For every IC $p=L[e]$ we first get its (generalized) faces that are before and after the IC on IP. For an IC inside the edge $e$
the function \emph{edge faces}, returns its adjacent faces $f_0$, $f_1$ (see Fig. \ref{fig:edge_faces}). 
Their order is given by the sign permuted inner products in 1d-2d intersection. The order of faces match 
the order of sides of IP if the sign is negative. If the sign is positive the function \emph{edge faces} returns face pair $(f_1, f_0)$.
\begin{figure}[!htb]
  \begin{center}        
    \includegraphics[width=0.3\textwidth]{\figpath edge_faces.pdf}
  \end{center}
  \caption{Order of faces adjacent to the oriented edge $e$ pointing up.}
  \label{fig:edge_faces}
\end{figure}

The function \emph{vertex faces} (Algorithm \ref{algo:vertex_faces}, described later)
returns a pair of generalized faces (face or edge) possibly adjacent to the IC $L[e]$ at the vertex $v$ of $S_3$.
On line \ref{line:mark_edges} we mark all edges coincident with the vertex, since if they have any other intersection corner they are degenerate and 
not processed anyway.

@@

@@

\begin{algorithm}
  \caption{2d-3d intersection, order the points}  
  \label{algo:order_23}

  \SetKw{To}{to}
  \DontPrintSemicolon
  \KwIn{all points in $I$}
  \KwOut{polygon in $K$ in correct order}

  \lIf{$\abs{I} < 3$}{ return $J$}
  \lElse{ return $I$ sorted according to connectivity in $F_g$ and $F_p$}
\end{algorithm}
\noteJB{Write correct algo with deduplication of vertices.}

@@

We see the benchmark results in the \fig{fig:fundamental_speed}. Green values correspond the presented algorithms
using \plucker coordinates, red values \notePE{***}. The gained speed up factor is approximately 5 in 1D-3D case and 2 in 2D-3D. Further we provide estimated count of floating point operations for different approaches in the Table \ref{tab:fundamental_flops}.

\subsection{General algorithms}
So far, we have suggested an intersection algorithm that uses axes aligned bounding boxes, BIH to find initial
intersection and advancing front method (shortly AFront) to prolong the intersection. From now on, we shall refer to an algorithm made of these components \emph{BIHsearch}. In this section, we shall compare two additional algorithms \emph{BBsearch} and \emph{BIHonly} to see the effects of BIH and AFront (see Table \ref{tab:search_algorithms}). \emph{BBsearch} does not use BIH, but searches for the initial intersection using only AABB, and then follows the AFront. The version 
\emph{BIHonly} computes AABB together with BIH, but does not use AFront.

@@

Next, we study the performance of the intersection algorithms on a mesh of a real problem, see \fig{fig:bedrichov_meshes}. The mesh represents a mountain ridge in the Jizera mountains which 
includes a system of geological fractures (\fig{fig:bedrichov_cut}). We also used this model to create
a less real mesh, in which the fractures extend through the bulk surface (\fig{fig:bedrichov_extend}). 
However, this situation might rather occur in another application than porous media problem.

>>>>

\section{Element Intersections}
\label{sec:element intersecitons}
In this section, we present algorithms for computing intersection of a pair of simplicial elements of a~different dimension in the 3D ambient space. 
In particular we are interested in the intersections of 1D-2D, 1D-3D, 2D-2D, 2D-3D pairs of elements. 
The fundamental idea is to compute intersection of 1D-2D simplices using the 
\plucker coordinates and reduce all other cases to this one.

@@

\subsection{Intersection Triangle-Tetrahedron (2D-2D)}
\noteJB{Just note that this algorithm can also be implemented using 1d-2d, but quite complex to not yet fully tested}

@@

\subsection{Intersection Triangle-Tetrahedron (2D-3D)}
The intersection of a triangle $S_2$ and a tetrahedron $S_3$ is an $n$-side polygon, $n\le 7$. The sides of the polygon
lie either on sides of $S_2$ or on faces of $S_3$. Thus each vertex (IC) of the polygon
can arise either from side-face intersection, or from edge-triangle intersection, or be a vertex of $S_2$.
So we have to compute at most 12 side-face intersections and at most 6 edge-triangle intersections. However,
to this end we only need to compute 9 \plucker coordinates (3 sides, 6 edges) and 18 permuted inner produts, one for every side-edge pair.
Computation of the intersection polygon consists of two parts: calculation of side-tetrahedron ICs (Algorithm \ref{algo:colect_23_ip_triangle})
provides all ICs on the boundary of $S_2$,
calculation of edge-triangle ICs (Algorithm \ref{algo:collect_23_ip_edges}).
\notePE{The following sentence is somehow weird.. I would say that we store ICs $I$; to sort $I$ during the computation, we use the two tables; at the end, $I$ is ordered correctly.}
Correct order of the ICs, stored in the list $I$, is defined by the \emph{connection tables} $F_g(:)$ and $F_p(:)$. \notePE{The following sentece is a copy of the second sentence is this subsection.} Every side of the intersection polygon 
lies either on a side of $S_2$, or on a face, or on an edge of $S_3$. Let us denote $M_2$ the set of sides of $S_2$ and $M_3$ the set of volume,
faces, edges, and vertices of $S_3$. Every side of polygon that lies on $x\in M_2\cup M_3$ is followed by an IC given by $F_g[x]$ 
and every IC $p$ is followed by the side that lies on $F_p[p]\in M_2\cup M_3$.

@@

  \DontPrintSemicolon
  \SetKw{continue}{continue}
  \KwIn{input data}
  \KwOut{List of ICs on sorted output data}  
  $F_g(:)=-1,\ F_p(:)=-1$\tcp*{Unset links.} 
  \For{side $s$ of $S_2$}{ 
    $L$ = intersection($s, S_3$)\;
    \lIf{ $\abs{L} = 0$}{\continue}
    \lIf{ $\abs{L} = 1$}{
        append $p$ to $J$
        \continue
    }
    \For{$p$ in $L$}{
        \lIf{$p \ne I[-1]$ }{append $p$ to $I$}
        \nllabel{line:m4_object}
        $p$ lies on $x\in M_3$\; 
        \uIf{$p$ is first in $L$}{
            $F_g[x] = p$,\ $F_p[p] = s$\;
        }
        \Else($p$ is the last in $L$){
            $F_g[s] = p,\ F_p[p] = x$\;
            \lIf{$F_g[x] = -1$}{$F_g[x] = p$} \nllabel{line:back_link43}
        }
    }
  }  
  \If{$I[-1] = I[0]$}{
    $x = F_p[I[-1]]$\;
    \If{$F_g[x]]=I[-1]$}{$F_g[x]=I[0]$}
    $F_g[s_2] = I[0]$\;
    remove $I[-1]$
  }
\end{algorithm}

Algorithm \ref{algo:colect_23_ip_triangle} passes through every side $s$ of the triangle $S_2$ and computes the line-tetrahedron intersection $L$.
In the regular case we process each of the two ICs in $L$. The IC $p$ is added to the list $I$ unless the last point is the same. 
This condition is effective just for the first IC in $L$ and merges ICs at the same vertex of $S_2$. Then we identify an object $x\in M_3$ the point
$p$ lies on (line \ref{line:m4_object}). \notePE{I am not sure, if the term 'object' is clear..}

\notePE{Why calling it vertex $v$ instead of IC $p$ here? What is $y$?}
If the vertex $v$ is between sides $s_1$, $s_2$, 
we effectively set connections $F_g[s_1] = v$, $F_g[y]=v$, $F_p[v] = s_2$. The backward temporary link on the line \ref{line:back_link43} 
is used in Algorithm \ref{algo:collect_23_ip_edges} to fix $F_p[p]$ for the second $p$ lying on an edge or a vertex of $S_3$. The condition at the end deals with the case
in which two sides intersect the same object $x\in M_3$.

For the regular intersection $L$, we pass through its ICs, as described above. \notePE{For special cases}, we firstly compute ICs
on the boundary of $S_2$ using the line-tetrahedron intersection algorithm for every side. We store single point intersections into separate list $J$
and skip filling the connection tables. This happens when the side touches $S_3$ at its edge or vertex. 
These ICs will be rediscovered again in Algorithm \ref{algo:collect_23_ip_edges} with better topological information, however
this is not the case if the touched edge $e$ of $S_3$ is coplanar with $S_2$ and the IC is inside of $e$. 
Therefore we keep a separate list $J$ of ICs to deal with this case.

@@

  \lFor{edge $e$ of $S_3$}{$L[e]$ = intersection($e, S_2$)}  \nllabel{line:12edges}
  \For{unmarked edge $e$ of $S_3$}{
    $p=L[e]$\;
    \uIf{ $p$ is inside $e$}{ 
        \nllabel{line:edge_faces}        
        $(f_0,\ f_1)$ = \edgefaces($e$)
    } 
    \Else($p$ at the vertex $v$ of $S_3$){       
            $(f_0,\ f_1)$ = \vertexfaces($v$,$L$)   \tcp*{Algorithm \ref{algo:vertex_faces}}
            \nllabel{line:mark_edges}
            mark all edges coincident with $p$
    }    
    \uIf{ $p$ is on boundary of $S_2$ }{
         $p$ lies on edge or at vertex $x \in M_3$\;
         $q=F_g[x]$\tcp*{$q$ is already computed $p$}
         \lIf{$F_p[q]=x$}{$F_p[q]=f_1$}
         \lElse{$F_g[f_0]=q$}
         $F_g[x]=-1$\tcp*{remove the backlink}
    }
    \Else{
         append $p$ to $I$\;
         $F_g[f_0]=p$\tcp*{overwrite the backlink}
         $F_p[p]=f_1$\;         
    }   
  }

  \lIf{$\abs{I} < 3$}{ return $J$}
  \lElse{ return $I$ sorted according to connectivity in $F_g$ and $F_p$}
\end{algorithm}

@@

Algorithm \ref{algo:collect_23_ip_edges} uses the line-triangle intersection algorithm for the edges of $S_3$ (line \ref{line:12edges}).
The loop produces ICs in the interior of $S_2$ and possibly those ICs with special position on vertex or edge of $S_3$ already computed in Algorithm \ref{algo:colect_23_ip_triangle}. 
Every edge $e$ of $S_3$ is oriented so that the pair of adjacent faces $f_0$, $f_1$ appears in the same order on the intersection polygon when the 
IC on $e$ has a negative sign (see Figure \ref{fig:plucker_products}). The function \emph{edge faces}, used on line \ref{line:edge_faces} and later on,
uses the sign of the intersection to return the pair of faces ordered correctly. 
Similarly the function \emph{vertex faces} (Algorithm \ref{algo:vertex_faces}, described later)
returns a pair of generalized faces (face or edge) possibly adjacent to the IC $L[e]$ at the vertex $v$ of $S_3$.
On line \ref{line:mark_edges} we mark all edges coincident with the vertex, since if they have any other intersection corner they are degenerate and 
not processed anyway.

@@

\begin{algorithm}
  \caption{2d-3d intersection, finish sort of points}  
  \label{algo:order_23}

  \SetKw{To}{to}
  \DontPrintSemicolon
  \KwIn{all points in $I$}
  \KwOut{polygon in $K$ in correct order}  
  $P=(1,\dots, n-1,0)$\tcp{array of sucessors}
  \lFor{ $f$ is face of $S_3$}{$P[ F[1,f] ] = F[0, f]$}
  $i=0$\;  
  \lFor{$n=0$ \To $\abs{I}-1$}{
    $K[n]=I[i]$
    $i=P[i]$
  }    
  \label{algo:order_23}
\end{algorithm}

@@

We see the benchmark results in the \fig{fig:fundamental_speed}. Green values correspond to the presented algorithms
using \plucker coordinates, red values \notePE{***}. The gained speed up factor is approximately 5 in 1D-3D case and 2 in 2D-3D. Further we provide estimated count of floating point operations for different approaches in the Table \ref{tab:fundamental_flops}.

\subsection{General algorithms}
So far, we have suggested an intersection algorithm that uses axes aligned bounding boxes, BIH to find initial
intersection and advancing front method (shortly AFront) to prolong the intersection. From now on, we shall refer to the algorithm made of these components \emph{BIHsearch}. In this section, we shall compare two additional algorithms \emph{BBsearch} and \emph{BIHonly} to see the effects of BIH and AFront (see Table \ref{tab:search_algorithms}). \emph{BBsearch} does not use BIH, but searches for the initial intersection using only AABB, and then follows AFront. The version 
\emph{BIHonly} computes AABB together with BIH, but does not use AFront.

@@

Next, we study the performance of the intersection algorithms on a mesh of a real problem, see \fig{fig:bedrichov_meshes}. The mesh represents a mountain ridge in the Jizera mountains which 
includes a system of geological fractures (\fig{fig:bedrichov_cut}). We also used this model to create
a less real mesh, in which the fractures extend through the bulk surface (\fig{fig:bedrichov_extend}). 
However, this situation might rather occur in another application than aporous media problem.
In total, 28 fractures are in the mesh.

The results for both meshes can be seen in the \fig{fig:bedrichov_speed}, pay attention to the different time
scales in the graph. In the first case, we notice that algorithms using AFront are nearly twice as fast as \emph{BIHonly}. The speed up corresponds to the time saved by using AFront instead of searching the whole mesh for candidate pairs. Computation of BIH in \emph{BIHsearch} pays off and the algorithm performs better than \emph{BBsearch}.

In the second case, we observe large blow up for \emph{BBsearch}. It is caused by the exterior elements, 
for which all bulk elements bounding boxes are iterated, so we can be sure there is no intersection.
This of course is much better performed using BIH.

In contrast to the artificial case on the fractured cube, BIH construction is worth due
to the presence of more components. Having more fractures inside the bulk would result to even bigger difference.
